
---
title: "Cyclistic Bike-Share Case Study"
author: "Udoh Chioma Lilian"
date: "2026-01-28"
output: html_document
---


## 1. Business Problem

Cyclistic, a bike-share company in Chicago, aims to increase the number of annual memberships, as membership growth is essential for the company’s long-term profitability and sustainability. However, to achieve this goal, the company must first understand how casual riders(customers who purchase single-ride or full-day passes) and annual members(customers who subscribe annually) use Cyclistic bikes differently.

## 2. Business Task
As a junior data analyst working with the marketing analytics team at Cyclistic, the objective of this project is to analyze user behaviour and identify behavioral differences between casual riders and annual members. The goal is to help the marketing team develop effective strategies to drive conversion. 

The final recommendations will be supported by clear data analysis and professional visualizations to ensure they are actionable and persuasive enough to gain approval from Cyclistic executives.



## 3. Data Source Description and Credibility

This case study uses [Divvy 2019 Q1](https://docs.google.com/spreadsheets/u/0/d/1HyGxbpi5Xad5ARKYb5CQL57LrhR2t_KwCt1CRXus5As/edit) and [Divvy 2020 Q1](https://docs.google.com/spreadsheets/u/0/d/1Y92ILGP-9Wyq418FYuVt14smYzt33G3-T8SoJMCo_UM/edit) datasets, provided by Lyft Bikes and Scooters, LLC (“bikeshare”), a bike-share company operating in Chicago. The data is made available under a public [license](https://divvybikes.com/data-license-agreement) by Motivate International Inc., granting permission to access, analyze, and use the data for lawful purposes.   

The datasets are structured in Excel format with rows representing individual trips and columns describing trip attributes. For the purpose of this case study, the datasets are appropriate, reliable and comprehensive, containing all necessary variables required for this analysis. Although the data is not fully current, it is appropriate for identifying usage patterns and behavioral differences between rider types.

Personally identifiable information has been removed, ensuring rider privacy. The data was handled securely on a password-protected device and used strictly for analytical purposes.

## 4. Data Cleaning and Preparation

```{r setup, include=FALSE}
library(dplyr)
library(stringr)
library(janitor)
library(lubridate)
library(ggplot2)
library(DT)
library(scales)
library(tidyverse)
library(readxl)
```

```{r load-data, include=FALSE}

# Load the 2019 data
Divvy_Trips_2019_Q1 <- read_excel("C:/Users/HP/Downloads/Divvy_Trips_2019_Q1.xlsx")

# Load the 2020 data
Divvy_Trips_2020_Q1 <- read_excel("C:/Users/HP/Downloads/Divvy_Trips_2020_Q1.xlsx")

```

```{r data-cleaning, include=FALSE}

  #To check for duplicate trip_id
  Divvy_Trips_2019_Q1 %>%
    count(trip_id) %>%
    filter(n > 1)
  #There are no duplicates
  
  
  #To ensure no spaces in column names and all column names are in lowercase
  clean_names(Divvy_Trips_2019_Q1) 
  
  
  #To view the new cleaned dataframe
  clean1_Divvy_Trips_2019_Q1 <- clean_names(Divvy_Trips_2019_Q1)
  
  
  #To count the number of rows with empty cells
  Divvy_Trips_2019_Q1 %>%
    filter(if_any(everything(), is.na)) %>%
    nrow()
  #There are 19712 rows
  
  
  #To delete empty cells
  clean2_Divvy_Trips_2019_Q1 <- drop_na(clean1_Divvy_Trips_2019_Q1)
  
  
  # To detect inconsistent suffixes using the full name of observed abreviations prominent in the dataframe
  patterns <- c("\\bRoad\\b", "\\bStreet\\b", "\\bParkway\\b", "\\bAvenue\\b",
                "\\bBoulevard\\b", "\\bPlace\\b", "\\bDrive\\b")
  
  
  # Combine into one regex
  regex_pattern <- paste(patterns, collapse = "|")
  
  
  # Filter rows where either start or end station has any of these exact patterns
  clean3_Divvy_Trips_2019_Q1 <- clean2_Divvy_Trips_2019_Q1 %>%
    filter(
      str_detect(from_station_name, regex_pattern) |
        str_detect(to_station_name, regex_pattern)
    )
  #Inconsistency detected in place appearing as "pl" in most places and "place" in a certain station name, "Mccormick Place"
  
  
  #Make the naming format consistent as "Pl" using global substitution
  clean4_Divvy_Trips_2019_Q1 <- clean2_Divvy_Trips_2019_Q1
  clean4_Divvy_Trips_2019_Q1$from_station_name <- 
  gsub("Place", "Pl", clean2_Divvy_Trips_2019_Q1$from_station_name)
  
  
  clean5_Divvy_Trips_2019_Q1 <- clean4_Divvy_Trips_2019_Q1
  clean5_Divvy_Trips_2019_Q1$to_station_name <-
  gsub("Place", "Pl", clean4_Divvy_Trips_2019_Q1$to_station_name)
  
  
  #To confirm fixed inconsistent suffixes
  patterns <- c("\\bPlace\\b")
  
  
  # Combine into one regex
  regex_pattern <- paste(patterns, collapse = "|")
  
  
  # Filter rows where either start or end station has any of these exact patterns
  clean5_Divvy_Trips_2019_Q1 %>%
    filter(
      str_detect(from_station_name, regex_pattern) |
        str_detect(to_station_name, regex_pattern)
    ) %>%
    select(from_station_name, to_station_name)
  #result is empty meaning it has been consistently formatted.
  
  
  clean6_Divvy_Trips_2019_Q1 <- clean5_Divvy_Trips_2019_Q1 %>%
    mutate(across(where(is.character), ~ str_to_title(str_squish(trimws(str_remove_all(., "\\(.*?\\)"))))))
  #To remove any leading or trailing spaces and remove the Asterisk appearing in some street address across all addresses.
  #library(dplyr) loads dplyr, which gives you functions like mutate(), across(), %>% (the pipe).
  #library(stringr) loads stringr, which has handy string functions like str_squish().
  #cleaned_Divvy_Trips_2019_data %>% ... the %>%  takes what you have written in the first code line and calls information from it with the next one
  #mutate(...) is saying “create new columns or change existing ones.”
  #across(where(is.character), ~ str_squish(trimws(.)))
  #across lets you apply a function to multiple columns at once rather than having to call out each columns one after the other.
  #where(is.character) picks all columns that are text/strings.
  #~ str_squish(...) implies for each cell in those columns:
  #str_to_title sets texts to title case
    #trimws(.) removes spaces at the start or end of the text
  #str_squish(.)removes extra spaces inside the text, leaving just one space between words
  
  
  #Rename from_station_name to start_station_name
  #Rename to_station_name to end_station_name
  #Rename from_station_id to start_station_id
  #Rename to_station_name to end_station_id
  
  
  clean6_Divvy_Trips_2019_Q1 <- clean6_Divvy_Trips_2019_Q1 %>%
    rename(start_station_name = from_station_name, 
           end_station_name = to_station_name,
           start_station_id = from_station_id, 
           end_station_id = to_station_id)
  
  #Longitude and latitude values were validated by confirming that latitude values
  # fell within the expected range (−90 to 90) and longitude values within (−180 to 180).
  # The observed negative longitude values are consistent with the study area being 
  #located west of the Prime Meridian.”
  clean7_Divvy_Trips_2019_Q1 <- clean6_Divvy_Trips_2019_Q1%>%
    select(-gender,-birthyear, -bikeid)

  
  #To confirm that trip_duration is stored in seconds
  tripduration_check <- clean7_Divvy_Trips_2019_Q1 %>%
    mutate(tripduration_seconds = as.numeric(difftime(end_time, start_time, units = "secs")))
  
  
  #To confirm that their calculation is correct
  tripduration_check <- tripduration_check %>%
    mutate(duration_diff = tripduration - tripduration_seconds)
  
  tripduration_check %>%
    filter(duration_diff != 0) %>%
    nrow()
  #16 diferences noted, Hence we use our calculated field
  
  
  clean8_Divvy_Trips_2019_Q1 <- tripduration_check%>%
    select(-tripduration, - duration_diff)
  
  
  
  
  #for divvy_trips_2020
  #To check for duplicate ride_id
  Divvy_Trips_2020_Q1 %>%
    count(ride_id)%>%
    filter(n>1)
  
  
  #To check through the dataframe and count how many rows have at least one empty cell
  Divvy_Trips_2020_Q1%>%
    filter(if_any(everything(), is.na))%>%
    nrow()
  
  
  #To delete empty cells
  clean1_Divvy_Trips_2020_Q1 <- drop_na(Divvy_Trips_2020_Q1)
  
  
  #To remove any leading or trailing spaces and remove the observed Asterisk appearing in some street address across all addresses.
  clean2_Divvy_Trips_2020_Q1 <- clean1_Divvy_Trips_2020_Q1%>%
    mutate(across(where(is.character), ~str_to_title(str_squish(trimws(str_remove_all(., "\\(.*?\\)"))))))
  
  
  #To ensure that there are only characters, letters and underscores in the dataframe
  clean3_Divvy_Trips_2020_Q1 <- clean_names(clean2_Divvy_Trips_2020_Q1)
  
  
  
  # To detect inconsistent suffixes using the full name of observed abreviations prominent in the dataframe
  patterns <- c("\\bRoad\\b", "\\bStreet\\b", "\\bParkway\\b", "\\bAvenue\\b",
                "\\bBoulevard\\b", "\\bPlace\\b", "\\bDrive\\b")
  
  # Combine into one regex
  regex_pattern <- paste(patterns, collapse = "|")
  
  # Filter rows where either start or end station has any of these exact patterns
  clean3_Divvy_Trips_2020_Q1 %>%
    filter(
      str_detect(start_station_name, regex_pattern) |
        str_detect(end_station_name, regex_pattern)
    ) %>%
    select(start_station_name, end_station_name)
  
  
  #This shows that place was formatted inconsistently as pl in most places and as place in some others. The others didn't appear in the dataframe
  #This shows 565 rows
  
  
  #Rectify this by using global substitution, gsub()
  
  
  clean4_Divvy_Trips_2020_Q1 <- clean3_Divvy_Trips_2020_Q1
  clean4_Divvy_Trips_2020_Q1$start_station_name <-
    gsub("Place", "Pl", clean3_Divvy_Trips_2020_Q1$start_station_name)
  
  clean5_Divvy_Trips_2020_Q1 <- clean4_Divvy_Trips_2020_Q1
  clean5_Divvy_Trips_2020_Q1$end_station_name <-
    gsub("Place", "Pl", clean3_Divvy_Trips_2020_Q1$end_station_name)
  
  #To confirm fixed inconsistent suffixes
  patterns <- c("\\bPlace\\b")
  
  
  # Combine into one regex
  regex_pattern <- paste(patterns, collapse = "|")
  
  
  # Filter rows where either start or end station has any of these exact patterns
  clean5_Divvy_Trips_2020_Q1 %>%
    filter(
      str_detect(start_station_name, regex_pattern) |
        str_detect(end_station_name, regex_pattern)
    ) %>%
    select(start_station_name, end_station_name)
  #result is empty meaning it has been consistently formatted.
  
  #Longitude and latitude values were validated by confirming that latitude values
  # fell within the expected range (−90 to 90) and longitude values within (−180 to 180).
  # The observed negative longitude values are consistent with the study area being 
  #located west of the Prime Meridian.”
  
  #Rename started_at to start_time and ended_at to end_time
  
  clean6_Divvy_Trips_2020_Q1 <- clean5_Divvy_Trips_2020_Q1 %>%
    rename(start_time = started_at, 
           end_time = ended_at,
           usertype = member_casual,
           trip_id = ride_id)
  
  #Remove columns for merging purpose
  clean7_Divvy_Trips_2020_Q1 <- clean6_Divvy_Trips_2020_Q1%>%
    select(-rideable_type, -start_lat, -start_lng, -end_lat, -end_lng)
  
  clean8_Divvy_Trips_2020_Q1 <- clean7_Divvy_Trips_2020_Q1 %>%
    mutate(tripduration_seconds = as.numeric(difftime(end_time, start_time, units = "secs")))
  
  
  #To make both datatypes character_datatype 
  clean8_Divvy_Trips_2019_Q1 <- clean8_Divvy_Trips_2019_Q1 %>%
    mutate(trip_id = as.character(trip_id))
  
  clean8_Divvy_Trips_2020_Q1 <- clean8_Divvy_Trips_2020_Q1 %>%
    mutate(trip_id = as.character(trip_id))
  
  
  merged_divvy_data1 <-
  bind_rows(clean8_Divvy_Trips_2019_Q1, clean8_Divvy_Trips_2020_Q1)
  
  
  #To change name of cell from customer to casual and from subscribers to members
  merged_divvy_data2 <- merged_divvy_data1 %>%
    mutate(usertype = recode(usertype,
                             "Customer" = "Casual",
                             "Subscriber" = "Member"))
  
  
  #To rearrange columns
  merged_divvy_data3 <- merged_divvy_data2%>%
    relocate(tripduration_seconds, .before = start_station_id)
  
  
  #Add day of week for the start_time and end_time
  
  merged_divvy_data4 <- merged_divvy_data3 %>%
    mutate(
      start_day_of_the_week = wday(start_time, label = TRUE),
      end_day_of_the_week   = wday(end_time,   label = TRUE)
    )
  
  
  #I arranged rows in descending order to check for false tripduration_seconds 
  #and spotted some -ve values and some values less than 60s which i use as my benchmark for rides
  #I also noticed some data with outrageous amounts of time spent on rides when u arranged by clicking on the 
  #arrow in the r source and set a benchmark to remove all greater than 24 hrs that is 86400 seconds
  
  Original_merged_rows <- nrow(merged_divvy_data4)
  
  
  merged_divvy_data5 <- merged_divvy_data4 %>%
    filter(tripduration_seconds > 0) %>%   # remove negative durations
    filter(tripduration_seconds >= 60) %>%    # remove very short trips
    filter(tripduration_seconds <= 86400)
  
  
  cleaned_merged_rows <- nrow(merged_divvy_data5)
  
  removed_merged_rows <- Original_merged_rows - cleaned_merged_rows
  
  removed_percent <- (removed_merged_rows / Original_merged_rows) * 100
  #1.048% of the data  
  
  #Rearranged datasets again
  merged_divvy_data6 <- merged_divvy_data5%>%
    relocate(start_day_of_the_week, end_day_of_the_week, .before = start_station_id)
  
  
  #ANALYSES PHASE
  
  #mean and median trip duration for each usertypes
  summary_tripduration <- merged_divvy_data6 %>%
    group_by(usertype) %>%
    summarise(
      avg_tripduration_secs   = mean(tripduration_seconds),
      median_tripduration_secs = median(tripduration_seconds)
    )
  #in minutes
  summary_tripduration %>%
    mutate(avg_tripduration_mins = avg_tripduration_secs / 60,
           median_tripduration_mins = median_tripduration_secs / 60)%>%
    select(-avg_tripduration_secs, -median_tripduration_secs)
  
#The avg_trip duration for casual users was 39 minutes and from member was 11 minutes
#However, the median trip duration is 22.3 minutes for casual users and 8.5 minutes for members.
#This implies that the distribution of trip durations is moderately positively skewed.
#, and this results from the presence of longer trips that increase the overall average, which is typical in bike-share usage data.
#This also implies that casual riders use bike for a longer duration than member.
  
  
  #We now dive deeper to observe how these two users differs by days of the week(dow)
  #I realized I still had two columns and decided to work with a single column for date but then 
  #going through my data I realised that some hours crossed over from one day into another and so
  #I picked the start day as my day of week and proceeded to exploring the differment between users by dow
  
  
  all(merged_divvy_data6$start_day_of_the_week ==
        merged_divvy_data6$end_day_of_the_week)
  
  merged_divvy_data6 %>%
    filter(start_day_of_the_week != end_day_of_the_week) %>%
    nrow()
  
  merged_divvy_data6 %>%
    filter(start_day_of_the_week != end_day_of_the_week) %>%
    select(start_time, end_time, start_day_of_the_week, end_day_of_the_week)
  
  
  #To remove end day of week and change start day to trip_day
  merged_divvy_data7 <- merged_divvy_data6 %>%
    select(- end_day_of_the_week)%>%
    rename(trip_day = start_day_of_the_week )

 
  
  daily_summary_member <- merged_divvy_data7 %>%
    filter(usertype == "Member") %>%
    group_by(trip_day) %>%
    summarise(
      avg_tripduration_min = mean(tripduration_seconds)/60,
      median_tripduration_min = median(tripduration_seconds/60),
      .groups = "drop"
    )
  
  daily_summary <- merged_divvy_data7 %>%
    group_by(usertype, trip_day) %>%
    summarise(avg_tripduration = mean(tripduration_seconds) / 60)
  
  ggplot(daily_summary, aes(trip_day, avg_tripduration, fill = usertype)) +
    geom_col(position = "dodge") +
    labs(title = "Average Trip Duration by Day of Week",
         x = "Day of Week",
         y = "Average Trip Duration (minutes)")
  
  #Casual riders consistently take longer trips across all days, 
  #with peaks on weekends. Members show stable, shorter ride duration during weekdays,
  #reinforcing commuter behavior.
  #The highest average trip duration for both casual and members were on sundays. 
  #For both groups, the mean is greater than the median showing that the data is  slightly skewed.
  #Casual users consistently took longer trips than members, with average trip durations 
  #peaking on Sundays. Median values show a similar pattern, 
  #confirming that casual users’ rides are both longer and slightly more variable.
  #for Lets see if this will 
  #be same for the count per day to see which day bike is being used the most.
 #Add a column for day_type to calculate avg count by day of the week
  merged_divvy_data8 <- merged_divvy_data7 %>%
    mutate(
      day_type = ifelse(trip_day %in% c("Sat", "Sun"),
                        "Weekend", "Weekday"))%>%
    relocate(day_type, .after = trip_day) 

  
  #daily trip
  daily_trip_counts <- merged_divvy_data8 %>%
    group_by(usertype, day_type, trip_day) %>%
    summarise(
      trip_count = n(),
      .groups = "drop"
    )
 
  
  #avg trip count by day_type
avg_day_type_trip_count <- merged_divvy_data8 %>%
  group_by(usertype, trip_day, day_type) %>%
  summarise(daily_trip_count = n(), .groups = "drop") %>%
  group_by(usertype, day_type) %>%
  summarise(
    avg_daily_trip_count = round(mean(daily_trip_count), 0),
    .groups = "drop"
  )

  
  ggplot(daily_trip_counts,
         aes(x = trip_day,
             y = trip_count,
             fill = usertype)) +
    geom_col(position = "dodge") +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Daily Trip Count by Day of the Week",
      x = "Day of the Week",
      y = "Number of Trips",
      fill = "User Type"
    ) +
    theme_minimal()
  

  
  # Plot the average trip counts
  ggplot(avg_day_type_trip_count, aes(x = day_type, y = avg_daily_trip_count, fill = usertype)) +
    geom_col(position = "dodge") +
    labs(
      title = "Average Daily Trip Count by User Type: Weekday vs Weekend",
      x = "Day Type",
      y = "Average Number of Trips",
      fill = "User Type"
    ) +
    theme_minimal()
  
  
  combined_summary <- merged_divvy_data8 %>%
    group_by(usertype, day_type) %>%
    summarise(
      avg_trip_count = n() / n_distinct(trip_day),
      avg_trip_duration_min = mean(tripduration_seconds) / 60,
      .groups = "drop"
    )
  combined_summary
  
  #Rearranged the dataset for plotting purposes
  combined_long <- combined_summary %>%
    pivot_longer(
      cols = c(avg_trip_count, avg_trip_duration_min),
      names_to = "metric",
      values_to = "value"
    )
  
  
  #plot
  
  ggplot(combined_long,
         aes(x = day_type, y = value, fill = usertype)) +
    geom_col(position = "dodge") +
    facet_wrap(~ metric, scales = "free_y",
               labeller = as_labeller(c(
                 avg_trip_count = "Average Trip Count",
                 avg_trip_duration_min = "Average Trip Duration (Minutes)"
               ))) +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Trip Frequency and Duration by User Type and Day Type",
      x = "Day Type",
      y = "Value",
      fill = "User Type"
    ) +
    theme_minimal()
 #Casual riders use bikes mostly on weekends with the amount of time bikes 
  #where ordered on sat and sun as 15507 and 8582 times.There was a huge drop on weekdays 
  #Members use bikes at a high count rate on weekdays, even about 7 times more than the casual riders use it on weekends
  #On average no of bikes picked up by members on weekdays are 119046 bikes and 59059 on weekends
  #While avg  count of bikes casual users is  12044 on weekends and 5047 on weekdays
  #we observed that bikes are used mostly by members than by casual riders. 
  #While long trips are made by casual users but they do not use bikes as often as the members
  #Hence, members use it very often but take shorter trips.
  
  
  #We now dive deeper to see which hour is the most recorded hrs bikes are used
  
  merged_divvy_data9 <- merged_divvy_data8 %>%
    mutate(start_hour = hour(start_time)) %>%
    relocate(start_hour, .before = trip_day)

  
  #see trip count  by start hour
  trips_by_hour_overall <- merged_divvy_data9 %>%
    group_by(start_hour) %>%
    summarise(trip_count = n()) %>%
    arrange(start_hour)

  
  #Overall , most trips occurs between 7am - 8am and 4pm -5pm . The least occurs at 3am predict that the
  #we dive deeper to seeing if this is so for each user
  #To see if there's a  difference in hourly behaviour 
  
  
  #trip Count by usertype and start hour 
  
  hourly_trip_by_usertype <- merged_divvy_data9 %>%
    group_by(usertype, start_hour) %>%
    summarise(trip_count = n(), .groups = "drop") %>%
    arrange(usertype, desc(trip_count))
  #Casual users peak around 13–17 (1pm – 5pm), afternoon/early evening
  #casual riders rarely ride at 0-5(12am - 5am) that is at night / very early morning
  #Members peak around 7–8am and again around 4pm - 6pm, mostly daytime commuter hours, tho late morning 9am and midday rides 12pm - 3pm exists but are not as intense at commuting hours
  #members similar to causal riders, ride least at night from 0am - 4 am
  
  
  
  #we now see if there are stations that are popular for members and for casual riders
  # Top start stations overall
  top_start_stations <- merged_divvy_data9 %>%
    group_by(start_station_name) %>%        # Group by start station
    summarise(station_count = n()) %>%        # Count trips per station
    arrange(desc(station_count))              # Sort descending

  
  top_start_stations_usertype <- merged_divvy_data9 %>%
    group_by(usertype, start_station_name) %>% # Group by user type + start station
    summarise(station_count = n()) %>%
    slice_max(station_count, n = 10)%>%
    arrange(usertype, desc(station_count)) 
  
  #we now see if theres stations that are popular for members and for casual riders
  # Top start stations overall
  top_end_stations <- merged_divvy_data9 %>%
    group_by(end_station_name) %>%        # Group by start station
    summarise(station_count = n()) %>%        # Count trips per station
    arrange(desc(station_count))              # Sort descending
 
  
  
  top_end_stations_usertype <- merged_divvy_data9 %>%
    group_by(usertype, end_station_name) %>% # Group by user type + start station
    summarise(station_count = n()) %>%
    slice_max(station_count, n = 10)%>%
    arrange(usertype, desc(station_count)) 
```


### 4.1 Cleaning Divvy 2019 Q1 Data

cleaning steps included:

* Checking and confirming absence of duplicate `trip_id` values
* Standardizing column names to lowercase with no spaces
* Removing rows with missing values
* Resolving inconsistent station name suffixes (e.g., "Place" to "Pl")
* Removing leading and trailing whitespace across all entries
* Removing asterisks present in some street address entries across all addresses
* Renaming columns for consistency with 2020 data
* Validating latitude and longitude ranges
* Recalculating trip duration in seconds using start and end times

Trips with incorrect or inconsistent duration values were corrected using calculated durations.

### 4.2 Cleaning Divvy 2020 Q1 Data

Cleaning steps mirrored the 2019 process and additionally included:

* Renaming columns to align with the 2019 schema
* Removing unnecessary columns to enable merging
* Creating a calculated trip duration field

### 4.3 Data Merging and Filtering

The cleaned 2019 and 2020 datasets were merged into a single dataset. User types were standardized as "Casual" and "Member". Additional cleaning included:

* Creating start and end day of the week columns
* Rearranging tripduration_seconds field to identify any anomalies that could exist
* Removing negative trip duration
* Excluding trips shorter than 60 seconds, as they are likely not genuine bike trips
* Removing trips longer than 24 hours (86,400 seconds), as they likely represent data errors or bike maintenance trips, not typical rider behavior.

Only 1.048% of the total data was removed, ensuring minimal data loss while improving data quality. I now rearranged my dataset so that related fields are positioned close together for better readability and logical flow. Below is a sample of the cleaned dataset:

```{r merged_divvy_data6, echo=FALSE, message=FALSE, warning=FALSE}
# Show structure of cleaned dataset

datatable(
  head(merged_divvy_data6, 10),
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    autoWidth = TRUE
  ),
  caption = "Sample of Cleaned and Merged Dataset"
)
```

## 5. Analysis

### 5.1 Comparing Average and Median Trip Durations by User Type


```{r trip-duration-summary, echo=FALSE}
library(DT)


summary_tripduration <- merged_divvy_data6 %>%
  group_by(usertype) %>%
  summarise(
    avg_tripduration_mins = round(mean(tripduration_seconds) / 60, 1), 
    median_tripduration_mins = round(median(tripduration_seconds) / 60, 1)
  )

# Interactive table
datatable(summary_tripduration, 
          options = list(pageLength = 5, dom = 't'), 
          caption = "Trip Duration Summary by User Type")
```

**Insight:** Casual riders take significantly longer trips (average ≈ 39 minutes) compared to members (average ≈ 11 minutes), which is reflected in the median values of 22.3 minutes for casual riders and 8.5 minutes for members. However, the distribution of trip duration is moderately positively skewed, and this results from the presence of longer trips that increase the overall average, which is typical in bike-share usage data. Next, we examine how their riding patterns vary across the days of the week.


### 5.2 Comparing Average Trip Duration by Day of the Week and User Type

```{r, echo=FALSE}

daily_summary <- merged_divvy_data7 %>%        
  group_by(usertype, trip_day) %>%
  summarise(avg_tripduration = mean(tripduration_seconds) / 60, .groups = "drop")
ggplot(daily_summary, aes(trip_day, avg_tripduration, fill = usertype)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = round(avg_tripduration, 1)), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, size = 3) +  # adds values on top
  labs(title = "Average Trip Duration by Day of Week",
       x = "Day of Week",
       y = "Average Trip Duration (minutes)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 24, face = "bold"),
    axis.title = element_text(size = 20),
    axis.text = element_text(size = 16),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 16)
  )
```

**Insight:** Casual riders consistently take longer trips across all days, with peaks on weekends. These notable peaks could suggest that they are leisure-oriented. Members show stable, shorter ride duration during weekdays, similar to the behavior of commuters. The highest average trip duration for both casual riders and members was on Sundays. Furthermore, we can gain more insight by examining the daily trip counts to see which daytype(weekdays and weekends) bikes are used the most. This will help guide our analysis toward strategies for converting casual riders into members.

### 5.3 Comparing Daily Trip Counts by User Type and Day of the Week

```{r daily_trip_count, echo=FALSE}
ggplot(daily_trip_counts,
       aes(x = trip_day,
           y = trip_count,
           fill = usertype)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_text(aes(label = comma(trip_count)),
            position = position_dodge(width = 0.9),
            vjust = -0.5,
            size = 3) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Daily Trip Count by Day of the Week",
    x = "Day of the Week",
    y = "Number of Trips",
    fill = "User Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 24, face = "bold"),
    axis.title = element_text(size = 20),
    axis.text = element_text(size = 16),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 16)
  )

```

**Insight:**This clearly shows that casual users use bikes mainly during weekends and members use it mainly during weekdays, and overall use bikes the most.


To make a sense of it all we see a graph showing the average daytype trip count and the average trip duration by daytype to see the difference in count of trips and duration of trips between the two groups.


```{r trip frequency and duration, echo=FALSE}
 #plot
ggplot(combined_long,
       aes(x = day_type, y = value, fill = usertype)) +
  
  geom_col(position = position_dodge(width = 0.9)) +
  
  geom_text(aes(label = comma(round(value, 0))),
            position = position_dodge(width = 0.9),
            vjust = -0.3,
            size = 4) +
  
  facet_wrap(~ metric, scales = "free_y",
             labeller = as_labeller(c(
               avg_trip_count = "Average Trip Count",
               avg_trip_duration_min = "Average Trip Duration (Minutes)"
             ))) +
  
  scale_y_continuous(labels = comma) +
  
  labs(
    title = "Trip Frequency and Duration by User Type and Day Type",
    x = "Day Type",
    y = "Value",
    fill = "User Type"
  ) +
  
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

**Insight:** For casual users, the average number of trips is about 12,238 on weekends and 5,150 on weekdays. Members, on the other hand, use bikes at a much higher rate on weekdays, recording about seven times more trips than casual riders do on weekends.Their average number of trips is approximately 119,137 trips on weekdays and 59,119 trips on weekends.
This comparison highlights a clear behavioral difference between casual riders and members. Members record significantly higher trip counts, particularly on weekdays, suggesting frequent and routine usage consistent with commuting behavior. In contrast, casual riders show lower trip frequencies, especially on weekdays, but exhibit longer average trip durations, particularly during weekends. This indicates that while casual riders use the service less often, they tend to engage in longer leisure-oriented rides when they do use the bikes.
 
```{r, echo = FALSE}
day_type_count_by_usertype <- merged_divvy_data9 %>%
    group_by(usertype, day_type) %>%
    summarise(
      trip_count = n(),
      .groups = "drop"
    )
trip_distribution_by_day_type <- day_type_count_by_usertype %>%
  group_by(day_type) %>%
  mutate(prop = trip_count / sum(trip_count)) %>%
  ungroup()

# Plot
ggplot(trip_distribution_by_day_type,
       aes(x = day_type,
           y = prop,
           fill = usertype)) +
  
  geom_bar(stat = "identity", position = "fill") +
  
  geom_text(aes(label = percent(prop, accuracy = 0.1)),
            position = position_fill(vjust = 0.5),
            color = "white",
            size = 5) +
  
  scale_y_continuous(labels = percent_format()) +
  
  
  
  labs(
    title = "User Type Distribution by Day Type",
    x = "Day Type",
    y = "Percentage of Trips",
    fill = "User Type"
  ) +
  
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )
```

**Insight:** Casual users contribute 4.1% of total weekday trip counts, while members account for 95.9%. On weekends, casual users contribute 17.2% of total trip counts, while members account for 82.8%.


```{r, echo = FALSE }

trip_distribution_by_usertype <- merged_divvy_data9 %>% group_by(usertype) %>% summarise(trip_count = n(), .groups = "drop") %>% mutate(percentage_total = round(trip_count / sum(trip_count) * 100, 1)) %>% ungroup()
ggplot(trip_distribution_by_usertype,
       aes(x = percentage_total,
           y = usertype,
           fill = usertype)) +
  
  geom_col(width = 0.6) +
  
  geom_text(aes(x = percentage_total / 2,
                label = paste0(percentage_total, "%")),
            color = "white",
            size = 4.5) +

  annotate("text",
           x = 50, 
           y = 1.0,
           label = "Casual users represent 6.6% of total trips,\nwhile Members account for 93.4%.",
           size = 5,
           hjust = 0.5) +

  labs(
    title = "Overall Trip Share by User Type",
    x = "Percentage of Total Trips",
    y = "User Type"
  ) +
  
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )
```

**Insight:** In general, casual users have a higher average trip duration than members. While this might initially seem beneficial, the financial team has evaluated other factors that affect total revenue. Longer trips do not necessarily translate into higher overall earnings. The company may instead prioritize consistent engagement, stable revenue, and long-term customer value. This is reflected in the data, as members contribute 93.4% of the total trip count, indicating that they drive the majority of system usage.

Next, we explore to see which time of the day the bikes are used the most to further see the behavioural difference between this users.

### 5.4 Comparing Hourly Ride Patterns by User Type

```{r hourly-patterns, echo=FALSE}
hourly_trips <- merged_divvy_data9 %>%
  group_by(usertype, start_hour) %>%
  summarise(trip_count = n(), .groups = "drop")

ggplot(hourly_trips, aes(start_hour, trip_count, color = usertype)) +
  geom_line() +
  labs(title = "Hourly Ride Patterns by Usertype",
       x = "Hour of Day",
       y = "Trip Count") +
  theme(
    plot.title = element_text(size = 24, face = "bold"),
    axis.title = element_text(size = 20),
    axis.text = element_text(size = 16),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 16)
  )
```

**Insight:** Members peak during morning and evening commute hours (7am – 9am, 4pm – 6pm), while casual riders peak in the afternoon, again, indicating leisure-based usage.

### 5.5 Popular Stations
Comparing Top Start Stations by User Type:

```{r top-stations, echo=FALSE}
top_start_stations_usertype %>%
  ggplot(aes(reorder(start_station_name, station_count), station_count, fill = usertype)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top Start Stations by User Type",
       x = "Station",
       y = "Trip Count") +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 14)
  )
```

**Insight:** Members tend to start trips from central, commuter-friendly locations, while casual riders favor recreational or tourist-heavy stations.


## 6. Key Findings

* Casual riders take longer but fewer trips, particularly on weekends
* Members ride more frequently with shorter durations, especially on weekdays
* Members’ usage patterns align with commuting behaviour
* Casual riders’ behavior aligns with leisure and recreational use

## 7. Recommendations

### *Personalised and simplified data communication*
Cyclistic can analyse riders’ historical trip data to provide simple summaries showing:

- How long casual riders typically ride  
- How much they spend on longer trips  
- How this compares with what they would pay under a membership  

Presenting this information in a clear and non-technical format can help casual riders better understand their own usage patterns. Cyclistic can use:

- In-app notifications  
- App pop-ups after a completed ride  
- Email summaries  

These methods are cheaper, less aggressive, and more likely to be read at the right time.

#### Pros:

- Helps riders make informed decisions about membership
- Low cost and scalable communication method
- Timely notifications are more likely to be read and acted upon

#### Cons:

- Risk of overloading users with notifications, causing disengagement
- Some users may ignore messages entirely
- Requires accurate and up-to-date data processing


### *Encourage a one-month trial membership (not annual first)*
Since many casual riders believe they “don’t ride enough to sign up as an annual member”,  Cyclistic should offer a one-month membership trial through in-app pop-ups and email notifications, using messages like *“Still doubting? Try it for one month and see for yourself”*. This lets users test the value of membership themselves. 

After the one-month trial, Cyclistic can show users a comparison of:

- What they actually spent during the month  
- What they would have spent without membership  

#### Pros:

- Reduces hesitation for new members
- Lets users compare spending and decide based on real experience
- Builds trust and encourages long-term membership

#### Cons:

- Operational costs of managing trial memberships
- Messaging needs to be carefully timed to maximize uptake

### *Offer a discounted annual membership after the trial*
For riders who complete the trial period, Cyclistic can offer a discounted annual membership. 

#### Pros:

- Targets users who already experienced the value, increasing conversion likelihood
- Rewards loyal trial users, enhancing brand perception
- Discounted offer incentivizes commitment without pressure

#### Cons:

- Discount reduces short-term revenue per user
- Risk that some users may only wait for discounts instead of paying full price

### *Target casual riders at the right time and place*
Since casual riders mostly ride on weekends and during specific daytime hours, Cyclistic can promote membership through:

- In-app notifications during peak casual riding hours  
- Posters and QR codes, at high-traffic casual start stations  
- App notifications shown near ride start and after ride completion 

#### Pros:

- Reaches casual riders during peak usage hours, increasing message effectiveness
- Physical and digital presence reinforces brand visibility
- Increases likelihood of trial sign-ups and conversions

#### Cons:

- Marketing efforts may require extra resources (posters, QR codes, app scheduling)
- Messages may still be ignored if users are not receptive
- Over-targeting can feel intrusive if not carefully managed


## 8. Conclusion

Casual and member riders use Cyclistic bikes in fundamentally different ways. By leveraging these behavioral insights, Cyclistic can design targeted and realistic strategies to convert casual riders into loyal annual members, ultimately driving sustainable growth and revenue stability.

## 9. Limitations

- Data covers only Q1 2019 and Q1 2020; patterns may vary in other quarters
- Seasonal or weather effects are not captured
- No demographic data available

  


